---
title: "Introduction to RcppArrayFire"
author: "Ralf Stubner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RcppArrayFire}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`RcppArrayFire` provides an `Rcpp` wrapper for the [ArrayFire Library](http://www.arrayfire.com/), 
an open source library that can make use of GPUs and other accelerators via CUDA or OpenCL. 

# Basic usage

## Calculating $\pi$ by simulation

Let's look at the classical example of calculating $\pi$ via simulation. The 
basic idea is to generate a large number of random points within the unit 
square. An approximation for $\pi$ can then be calculated from the ratio of 
points within the unit circle to the total number of points. A vectorized 
implementation in R might look like this:

```{r, eval = FALSE}
piR <- function(N) {
    x <- runif(N)
    y <- runif(N)
    4 * sum(sqrt(x^2 + y^2) < 1.0) / N
}

set.seed(42)
system.time(cat("pi ~= ", piR(10^7), "\n"))
#> pi ~=  3.140899
#>        user      system     elapsed 
#>       0.836       0.060       0.897
```

A simple way to use C++ code in R is to use the inline package or `cppFunction()`
from Rcpp, which are both possible with RcppArrayFire. An implementation in C++
using ArrayFire might look like this:

```{r, eval = FALSE}
src <- '
double piAF (const int N) {
    array x = randu(N, f32);
    array y = randu(N, f32);
    return 4.0 * sum<float>(sqrt(x*x + y*y) < 1.0) / N;
}'
Rcpp::cppFunction(code = src, depends = "RcppArrayFire", includes = "using namespace af;")

RcppArrayFire::arrayfire_set_seed(42)
system.time(cat("pi ~= ", piAF(10^7), "\n"))
#> pi ~=  3.141066
#>        user      system     elapsed 
#>       0.000       0.004       0.021
```

Several things are worth noting:

(1) The syntax is almost identical. Besides the need for using types and a
different function name when generating random numbers, the argument `f32` to
`randu` as well as the `float` type catches the eye. These instruct ArrayFire to
use single precision floats, since not all devices support double precision
floating point numbers. If you want to use double precision, you have to specify
`f64` and `double`.

(2) The results are not the same, since ArrayFire uses a different random number
generator.

(3) The speed-up can be quite impressive. However, sometimes the first invocation of
a function is not as fast as expected due to the  [just-in-time compilation](http://arrayfire.com/performance-of-arrayfire-jit-code-generation/)
used by ArrayFire.


## Arrays as parameters 

Up to now we have only considered simple types like `double` or `int` as function
parameters and return values.  However, we can also use arrays. Consider the
matrix product $X^{\mathsf{T}} X$ for a random matrix $X$ in R:

```{r, eval = FALSE}
set.seed(42)
N <- 40
X <- matrix(rnorm(N * N * 2), ncol = N)
tXXR <- t(X) %*% X
```

The matrix multiplication can be implemented with RcppArrayFire using the
appropriate [matmul function](http://arrayfire.org/docs/group__blas__func__matmul.htm):

```{r, eval = FALSE}
src <- '
af::array squareMatrix(const RcppArrayFire::typed_array<f32>& x) {
    return af::matmulTN(x ,x);
}'
Rcpp::cppFunction(code = src, depends = "RcppArrayFire")
tXXGPU <- squareMatrix(X)

all.equal(tXXR, tXXGPU)
#> [1] "Mean relative difference: 1.372856e-07"
```

Since an object of type `af::array` can contain different [data types](http://arrayfire.org/docs/defines_8h.htm#a023d8ac325fb14f1712a52fb0940b1d5),
the templated wrapper class `RcppArrayFire::typed_array<>` is used to indicate
the desired data type when converting from R to C++. Again single precision
floats are used with ArrayFire, which explains the difference between the two
results. We can be sure that double precision is supported by switching the
computation backend to "CPU", which produces identical results:

```{r, eval = FALSE}
src <- '
af::array squareMatrixF64(const RcppArrayFire::typed_array<f64>& x) {
    return af::matmulTN(x ,x);
}'
Rcpp::cppFunction(code = src, depends = "RcppArrayFire")

RcppArrayFire::arrayfire_set_backend("CPU")
tXXCPU <- squareMatrixF64(X)
RcppArrayFire::arrayfire_set_backend("DEFAULT")

all.equal(tXXR, tXXCPU)
#> [1] TRUE
```


# Usage in a package

More serious functions should be defined in a permanent fashion. To facilitate
this, RcppArrayFire contains the function `RcppArraFire.package.skeleton()`.
This functions initialises a package with suitable `configure` script for
linking with ArrayFire and RcppArrayFire. In order to implement new
functionality you can then write C++ functions and save them in the `src`
folder. Functions that should be callable from R should be marked with the
`[[Rcpp::export]]` attribute. See the Rcpp vignettes on
[attributes](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-attributes.pdf) and
[package writing](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-package.pdf) for
further details.
